A Simple Type Theory for Category Theory

### 2022-09-01 paper/paper.tex
a table comparing with higher order logic

%TODO mandatory: add short abstract of the document
\begin{abstract}
  We present a domain-specific type theory for constructions and
  proofs in category theory. The type theory axiomatizes notions of
  category, functor, profunctor and a generalized form of natural
  transformations. By restricting the type theory to a form of ordered
  logic, we guarantee that all mappings between categories are
  functorial and all transformations are natural by construction, with
  no separate proof necessary. Important category theoretic proofs
  such as the Yoneda lemma and Co-yoneda lemma become simple type
  theoretic proofs about the relationship between unit, tensor and
  (ordered) function types. We demonstrate that this type theory can
  be used as a kind of logical framework in which we can define common
  categorical structures such as categories with families that serve
  as models of type theories.

@@

% TODO: do we have citations for this?
While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex, and there is difficulty in
acquisition of competence in category theory by computer scientists
who do not have a strong background in ``pure'' mathemantics.
%
While much of this can be explained by lack of, or lack of awareness
of, accessible material for computer scientists, in some ways category
theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and Kan extensions produce a proliferation of ``naturality''
and ``functoriality'' side-conditions that must be discharged.
%
This is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
These generalizations have long been recognized as being important in
programming language applications: domain-,metric- and
step-index-enriched categories have been used to model general
recursion and internal categories have been used to model
parametricity and gradual typing.

% TODO: Yoneda example
Novices in category theory are often overwhelmed by the number of
naturality and functoriality side conditions that abound in otherwise
simple proofs in category theory. For instance consider the proof of
the Yoneda lemma, often the first ``big'' proof in an introductory
category theory text. ... Here the reader is left to themselves to
show that the ``evaluation at identity'' morphism and its inverse are
actually natural transformations. They will then work through a
diagram chase to determine that this naturality condition indeed
holds.

Fortunately, the tools of category theory itself can be employed to
simplify these constructions, specifically the tools of \emph{higher}
category theory.
%
Consider the analogous situation in differential calculus. When a
physicist writes down a function, they do not expand out the
$\epsilon-\delta$ definition of continuity and prove that it is
continuous. Rather, they are aware of certain \emph{syntactic
principles}, most importantly that composition of continuous functions
is continuous and that previously defined functions are known to be
continuous.
%
Similar principles apply to category theory itself: a composition of
functors is functorial, a composition of natural transformations is
natual, etc.
%
In the case of category theory, these principles can be formalized
using category-like algebraic structures such as 2-categories,
bicategories, (virtual) double categories and pro-arrow
equipments\cite{}.
%
This approach has the advantage of avoiding the need to prove
the side conditions of naturality or enrichment.
%
The downside is that these algebraic structures are quite complicated,
with a 2-dimensional diagrammatic language that can be quite beautiful
and elegant, but is also seems very far removed from the traditional
formulation of category theory that while complex is at least
described in terms of a concrete notion of sets or types.

@@

@@

\subsection{Category of Elements}

One final multiplicative we can define is the \emph{category of
elements} of a profunctor. Given a profunctor $\alpha:\cat
C;\beta:\cat D \vdash R$, the ``category of elements'' of $R$ is a
category $\Sigma_{\alpha:\cat C;\beta:\cat D}. R$. In detail, its
objects are triples $(a:\cat C, r : R, b : \cat D)$ and morphism from
$(a,r,b)$ to $(a', r', b')$ consists of a pair of morphisms $g : b \to
b'$ and a backwards $f : a' \to a$ such that $r' = g\circ r \circ f$.
%
By picking $R = \punitinXfromYtoZ {\cat C} {\alpha}{\alpha'}$ we
recover the \emph{arrow} category $\exists
(\alpha;\alpha'). \punitinXfromYtoZ {\cat C}{\alpha}{\alpha'}$, by
composing these with functors, we recover the notion of \emph{comma}
category $\exists \alpha;\beta. \punitinXfromYtoZ {\cat
  E}{(F\alpha)}{(G\beta)}$ and by picking either $F$ or $G$ to have
domain the terminal category $1$, we recover the notion of a
\emph{slice} category.

This is unique among the multiplicatives in that it defines a
\emph{category} rather than a profunctor. Also note that while similar
to the tensor in that we use existential notation, the graph of a
profunctor is a \emph{negative} multiplicative, analogous to a
$\Sigma$-type. The universal property says that a functor from $\cat
E$ into the graph $\exists (\alpha; \beta). R$ is equivalent to a
transformation from $\cat E$ into the functor $R$, including functors
from $\cat E$ to $\cat C$ and $\cat D$.

@@

%% %% Please use bibtex,

\bibliography{lipics-v2021-sample-article}

>>>>

\begin{abstract}
  We present a domain-specific type theory for constructions and
  proofs in category theory. The type theory axiomatizes notions of
  category, functor, profunctor and a generalized form of natural
  transformations. By restricting the type theory to a form of ordered
  higher-order logic, we guarantee that all mappings between categories
  are functorial and all transformations are natural by construction,
  with no separate proof necessary. Important category theoretic
  proofs such as the Yoneda lemma and Co-yoneda lemma become simple
  type theoretic proofs about the relationship between unit, tensor
  and (ordered) function types.

@@

% TODO: do we have citations for this?
While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex, and there is difficulty in
acquisition of competence in category theory by computer scientists
who do not have a strong background in ``pure'' mathemantics.
%
While much of this can be explained by lack of, or lack of awareness
of, accessible material for computer scientists, in some ways category
theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and Kan extensions produce a proliferation of ``naturality''
and ``functoriality'' side-conditions that must be discharged.
%
This is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
These generalizations have long been recognized as being important in
programming language applications: domain-,metric- and
step-index-enriched categories have been used to model general
recursion and internal categories have been used to model
parametricity and gradual
typing\cite{order-enriched-cats,topos-of-trees,reflexive-graphs,double-cats-gradual-typing}.

% TODO: Yoneda example
Novices in category theory are often overwhelmed by the number of
naturality and functoriality side conditions that abound in otherwise
simple proofs in category theory. For instance consider the proof of
the Yoneda lemma, often the first ``big'' proof in an introductory
category theory text. The Yoneda lemma states that for any functor $P
: C^o \to \Set$ and object $A \in C$, that the set of natural
transformations $C(-,A) \to P$ is in bijection with the elements
$P(A)$. Furthermore, this bijection is natural in both $A$ and $P$. In
the authors' experience teaching category theory the sheer amount of
naturality side conditions in this result can be very intimidating to
learners, while experienced practitioners will observe that these
naturality conditions are tedious to carry out but routine. In our
type theory, none of these naturality conditions need to be explicitly
proven. Instead, all terms expressible in the type theory are
automatically natural by construction.

Fortunately, the tools of category theory itself can be employed to
simplify these constructions, specifically the tools of \emph{higher}
category theory.
%
Consider the analogous situation in differential calculus. When a
physicist writes down a function, they do not expand out the
$\epsilon-\delta$ definition of continuity and prove that it is
continuous. Rather, they are aware of certain \emph{syntactic
principles}, most importantly that composition of continuous functions
is continuous and that previously defined functions are known to be
continuous.
%
Similar principles apply to category theory itself: a composition of
functors is functorial, a composition of natural transformations is
natural, etc.
%
In the case of category theory, these principles can be formalized
using algebraic structures such as 2-categories, bicategories,
(virtual) double categories and pro-arrow equipments\cite{}.
%
This approach has the advantage of avoiding the need to prove
the side conditions of naturality or enrichment.
%
The downside is that these algebraic structures are quite complicated,
with a 2-dimensional diagrammatic language that can be quite beautiful
and elegant, but is also seems very far removed from the traditional
formulation of category theory that while complex is at least
described in terms of a concrete notion of sets or types.

@@

In Figure~\ref{fig:hol-cats} we give a table summarizing the
relationship between the judgments and connectives of higher-order
logic with our category logic. Since category logic is an ordered
variant of higher-order logic, some notions splinter into
multiple. For instance, covariant and contravariant presheaf
categories generalize the power set. Further, because we only have
binary relations rather than relations of arbitrary arity, we have
only restricted forms of universal and existential quantification
which combined with implications and conjunctions.

\begin{figure}
  \begin{center}
    \begin{tabular}{ |c|c| } 
      \hline
      Higher-Order Logic & Category Logic \\
      \hline
      Set $X$ & Category $\mathbb{C}$\\
      $X \times Y$ & $\mathbb C \times \mathbb D$ \\ 
      1 & $1$ \\
      $\mathcal P X$ & $\mathcal P^+ X, \mathcal P^- X$ \\
      $\{ (x,y) \in X \times Y | R(x,y) \}$ & $\sum_{\alpha:C;\beta:D} R$\\
      Function $f(x:X) : Y$ & Functor $F(\alpha:\cat C): \cat B$\\
      Relation $R(x,y)$ & Profunctor $R(\alpha:\cat C; \beta: \cat D)$\\
      $R \wedge Q$ & $R \times Q$\\
      $\top$ & $1$\\
      $\forall x. P \Rightarrow Q$ & $\homrallXYtoZ {\alpha:\cat C} P Q, \homlallXYtoZ {\alpha:\cat C} P Q$ \\
      $\exists x. P \wedge Q$ & $\tensorexistsXwithYandZ {\alpha:\cat C} P Q$\\
      $x =_X y$ & $\punitinXfromYtoZ {\cat C} \alpha \beta$\\
      Proof $R_1 \wedge \cdots \Rightarrow Q$ & Natural Transformation $x_1: R_1,\ldots \vdash t : Q$\\
      \hline
    \end{tabular}
  \end{center}
  \caption{Analogy between Higher-Order Logic and Category Logic
    Judgments and Connectives}
\end{figure}

@@

\subsection{Graph of a profunctor}

One final multiplicative we can define is the \emph{graph} of a
profunctor. Given a profunctor $\alpha:\cat C;\beta:\cat D \vdash R$,
the ``graph'' of $R$ is a category $\Sigma_{\alpha:\cat C;\beta:\cat
  D}. R$. In detail, its objects are triples $(a:\cat C, r : R, b :
\cat D)$ and morphism from $(a,r,b)$ to $(a', r', b')$ consists of a
pair of morphisms $g : b \to b'$ and a backwards $f : a' \to a$ such
that $r' = g\circ r \circ f$.
%
Intuitively this generalizes the ``graph'' of a relation $R : A\times
B \to \Omega$, where we think of the graph as the corresponding subset
$gr(R) \subseteq A\times B$
%
By picking $R = \punitinXfromYtoZ {\cat C} {\alpha}{\alpha'}$ we
recover the \emph{arrow} category $\exists
(\alpha;\alpha'). \punitinXfromYtoZ {\cat C}{\alpha}{\alpha'}$, by
composing these with functors, we recover the notion of \emph{comma}
category $\Sigma \alpha;\beta. \punitinXfromYtoZ {\cat
  E}{(F\alpha)}{(G\beta)}$ and by picking either $F$ or $G$ to have
domain the terminal category $1$, we recover the notion of a
\emph{slice} category.

This is unique among the multiplicatives in that it defines a
\emph{category} rather than a profunctor. The graph of a profunctor is
a \emph{negative} multiplicative, a kind of $\Sigma$-type for
profunctors. The universal property says that a functor from $\cat E$
into the graph $\Sigma (\alpha; \beta). R$ is equivalent to a
transformation from $\cat E$ into the functor $R$, including functors
from $\cat E$ to $\cat C$ and $\cat D$.
%
The graph of a profunctor is a joint generalization of several common
constructions in category theory: the arrow category, slice
categories, comma categories and the covariant and the covariant and
contravariant category of elements.

@@

%% %% Please use bibtex, 
\subsubsection*{Acknowledgements}
Grants stuff...? The authors would like to thank Emily Riehl, Mike
Shulman,...for helpful feedback on this work

\bibliography{cats}
